Depression, Anxiety, and Objective Hardship: Examining Prenatal Mental Health During the Covid-19 Pandemic in Canada

### 2024-04-07 outputs/paper.qmd
correct two grammar mistakes

Two new binary variables named `depression` and `anxiety` are created. Mothers scoring above 12, which represents 40% on the EPDS scale, are likely to be suffering from depression and should seek medical attention [@EDPSstanford]. Consequently, the variable `depression` is assigned to a value of 1 if the EPDS score is greater than or equal to 0.4, and 0 otherwise. Similarly, the `PROMIS` score of at least 20 indicates at least moderate depression. Hence, the variable `anxiety` is assigned to 1 if the `PROMIS` is larger or equal to 0.464, and 0 if it is smaller than 0.464.

@@

## Model Justification
For the depression model, I expect to see a positive relationship between experiencing at least moderate depression and age under 27, low levels of education (e.g. less than high school diploma), low income household in 2019 (e.g. less than 20,000), and preference of using English. This expectation is based on @sec-data. Moreover, at the onset of the pandemic, when everything was uncertain, it is reasonable to that participants with lower income would worry more about the pregnancy, because financial strain can contribute to limited access to resources such as healthcare. Younger individuals or those with lower levels of education may experience increased depression due to limited life experience or reduced access to information. Pregnant individuals who perceive a higher level of threat to themselves or their unborn baby may also experience heightened stress levels.

@@

## Birth Weight/Length Related to Prenatal Mental Health
Study has showed that maternal distress may contribute to reduced birthweight, and complications related to preterm birth and low birth weight are leading causes of infant morbidity and mortality [@NOWWUADCWWWD]. The dataset used in this paper, from [@osfPregnancyDuring], also demonstrate a reduced birth weight and birth length related to poor mental health. As illustrated in [@fig-weight_length_delivery], there is a clear association between higher levels of depression or anxiety and lower birth weight and length, categorized by delivery mode, with Caesarean-section babies having a larger average length and weight compared to vaginally delivered babies. These findings, again, emphasize the importance of addressing maternal mental health during pregnancy to support fetal growth and development.
```{r}
#| label: fig-weight_length_delivery
#| fig-cap: Distribution of Birth Weight and Length and Prenatal Mental Health by Delivery Mode
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| fig-subcap: ["Birth Weight and Depression by Delivery Mode", "Birth Weight and Anxiety by Delivery Mode", "Birth Length and Depression by Delivery Mode", "Birth Length and Anxiety by Delivery Mode"]
clean_data |>
  ggplot(aes(x = EPDS, y = birth_weight, color = delivery_mode)) +
  geom_smooth(method = "lm")+
  theme_minimal() + labs(x = "The Level of Prenatal Depression", y = "Birth Weight (grams)")

>>>>

Two new binary variables named `depression` and `anxiety` are created. Mothers scoring above 12, which represents 40% on the EPDS scale, are likely to be suffering from depression and should seek medical attention [@EDPSstanford]. Consequently, the variable `depression` is assigned to a value of 1 if the EPDS score is greater than or equal to 0.4, and 0 otherwise. The PROMIS scale scoring more than 20 indicates at least moderate depression. Hence, the variable `anxiety` is assigned to 1 if the PROMIS score is larger or equal to 0.464, and 0 if it is smaller than 0.464.

@@

## Model Justification
For the depression model, I expect to see a positive relationship between experiencing at least moderate depression and age under 27, low levels of education (e.g. less than high school diploma), low income household in 2019 (e.g. less than 20,000), and preference of using English. This expectation is based on @sec-data. Moreover, at the onset of the pandemic, when everything was uncertain, it is reasonable that participants with lower income would worry more about the pregnancy, because financial strain can contribute to limited access to resources such as healthcare. Younger individuals or those with lower levels of education may experience increased depression due to limited life experience or reduced access to information. Pregnant individuals who perceive a higher level of threat to themselves or their unborn baby may also experience heightened stress levels.

@@

## Birth Weight/Length Related to Prenatal Mental Health
Study has showed that maternal distress may contribute to reduced birthweight, and complications related to preterm birth and low birth weight are leading causes of infant morbidity and mortality [@NOWWUADCWWWD]. The dataset used in this paper, from the Open Science Frame [@osfPregnancyDuring], also demonstrate a reduced birth weight and birth length related to poor mental health. As illustrated in [@fig-weight_length_delivery], there is a clear association between higher levels of depression or anxiety and lower birth weight and length, categorized by delivery mode, with Caesarean-section babies having a larger average length and weight compared to vaginally delivered babies. These findings, again, emphasize the importance of addressing maternal mental health during pregnancy to support fetal growth and development.
```{r}
#| label: fig-weight_length_delivery
#| fig-cap: Distribution of Birth Weight and Length and Prenatal Mental Health by Delivery Mode
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| fig-subcap: ["Birth Weight and Depression by Delivery Mode", "Birth Weight and Anxiety by Delivery Mode", "Birth Length and Depression by Delivery Mode", "Birth Length and Anxiety by Delivery Mode"]
clean_data |>
  ggplot(aes(x = EPDS, y = birth_weight, color = delivery_mode)) +
  geom_smooth(method = "lm")+
  theme_minimal() + labs(x = "The Level of Prenatal Depression", y = "Birth Weight (grams)")
